---
title: "GGPlot"
author: "Data Science at Sac State"
date: "September 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GGPlot2

"ggplot2 is a data visualization package for the statistical programming language R. Created by Hadley Wickham in 2005, ggplot2 is an implementation of Leland Wilkinson's Grammar of Graphics—a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers." - <https://en.wikipedia.org/wiki/Ggplot2>

## Cheat Sheet

When starting to learn ggplot, remembering abbreviations can be tricky. Here's a cheat sheet with various functions and graphing types:

<https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf>.

## Setup

(Run install.packages('ggplot2') to install)

```{r library}
library(ggplot2)
```

## Basics

To make a graph, call 'ggplot()' and then append what you'd like to display (like layering components together). The simplest graph of this form follows, using a range of values along the X axis and Y axis.

```{r x_eq_y}
ggplot() + geom_point(aes(x = 0:10, y = 0:10))
```

We don't have to provide a value expression in the aes, we can also provide column names - so long as we provide the "data" argument to ggplot with the data frame containing our columns.

```{r}
# Mock data in a dataframe (same "ranges" as before)
df = data.frame(X = 0:10, Y = 0:10)

# ggplot will resolve the column names using the data frame
ggplot(data = df) +
  geom_point(aes(x = X, y = Y))
```

## A Non-Exhaustive List of Graph Types

### Blank
```{r}
ggplot() + geom_blank()
```

### Points
```{r}
ggplot() + geom_point(aes(x = 0:10, y = 0:10))
```

### Jitter
(Adds random jitter to points)

```{r}
ggplot() + geom_jitter(aes(x = 0:10, y = 0:10))
```

### Line
```{r}
ggplot() + geom_line(aes(x = 0:10, y = c(0, 3, 2, 1, 5, 2, 3, 5, 1, -2, 0)))
```

### Bar
```{r}
ggplot() + geom_bar(aes(x = c('a', 'b', 'b', 'b', 'c', 'c')))
```

### Histogram
Like bar, for continuous variables. Bin width can be specified.

```{r}
thousandRollsOfTenDice = replicate(1000, sum(sample(0:6, 10, replace=TRUE)))
ggplot() + geom_histogram(binwidth=2, aes(x = thousandRollsOfTenDice))
```

## Labels

Graph labels can be changed by appending the functions:

- ggtitle
- xlab
- ylab
- labs

#### Exercise

Make a blank graph with custom labels.

```{r echo=FALSE}
ggplot() +
  geom_blank() +
  ggtitle("Half Life 3 Game Reviews") +
  xlab("Time Since Game Release") +
  ylab("Rating Score")
```

## Faceting

Sometimes you want to make multiple graphs, each representing a different aspect of your data. You could write out a function that takes in the data and the desired value to filter out the data for, or you could "facet" and let ggplot do this for you.

### Example

Let's histogram the hours of sleep of various mammals.

```{r}
# msleep is a dataset built into ggplot. When you library(ggplot2) you import this data.
ggplot(data = msleep) +
  geom_histogram(binwidth=2, aes(x = sleep_total))
```

Now let's facet based this graph by what the mammals eat (the "vore" column of the data).

```{r}
# msleep is a dataset built into ggplot. When you library(ggplot2) you import this data.
ggplot(data = msleep) +
  geom_histogram(binwidth=2, aes(x = sleep_total)) +
  facet_wrap(~vore)
```

### Exercise

The 'diamonds' dataset is also built into ggplot. It contains pricing and sizing data of 50,000 diamonds. Columns include:

- Carat
- Cut
- Depth
- Price
- X/Y/Z (sizing info)

#### 1. Plot "x" size on X axis, Price on Y 

```{r echo=FALSE}
ggplot(data = diamonds) +
  geom_point(aes(x = x, y = price))
```

#### 2. Plot same graph as before, faceting on cut

```{r echo=FALSE}
ggplot(data = diamonds) +
  geom_point(aes(x = x, y = price)) +
  facet_wrap(~cut)
```

## Coloring

In addition to faceting, we can color the data based on a variable. Here's the same mammal sleep histogram as above, but colored by diet:

```{r}
# msleep is a dataset built into ggplot. When you library(ggplot2) you import this data.
ggplot(data = msleep) +
  geom_histogram(binwidth=2, aes(x = sleep_total, color = vore))
```

Note: To make the histogram fill the bars rather than outline, use the "fill" aes parameter instead.

#### Exercise

Redo the faceted diamond example, coloring based on carat. Because carat is a continuous variable, it will use a gradient rather than a distinct palatte.

```{r echo=FALSE}
ggplot(data = diamonds) +
  geom_point(aes(x = x, y = price, color = carat)) +
  facet_wrap(~cut)
```

## Alpha

In the last graph, the clustering of points makes it hard to tell how many points are layered over each other. If we make each point nearly transparent, we'll be able to see darker regions where there are more points.

Here's a graph showing translucent points:

```{r}
# rnorm provides N random values over normal distribution from -1 to 1
ggplot() +
  geom_point(aes(x = rnorm(5000), y = rnorm(5000)), alpha = 0.1)
```

### Exercise

Modify the last "diamond" graph (faceted or non-faceted) to make each point slightly transparent. Find a nice alpha value. Hint: Start with and alpha value of something like .1 so you can see when things are working.

```{r echo=FALSE}
ggplot(data = diamonds) +
  geom_point(aes(x = x, y = price), alpha = 0.009)
```

This helps us see the disparity of diamiond sizes.

## Point Sizing

You can change the size/shape of points with more aesthetics, here is a sizing example:

```{r}
# rnorm provides a random value over normal distribution
ggplot(data = data.frame(X = rnorm(200), Y = rnorm(200))) +
  geom_point(aes(x = X, y = Y, size = X+Y))
```

## FINAL EXAM

Make a graph using the iris data set.

- Position on X axis based on petal width
- Position on Y axis based on petal length
- Size based on Sepal Length + Sepal Width (it's as easy as you think)
- Color based on Species
- Relabel 

```{r echo=FALSE}
ggplot(data = iris) +
  geom_point(aes(x = Petal.Width, y = Petal.Length, size = Sepal.Length + Sepal.Width, color = Species)) +
  ggtitle("Iris Sizes by Species") +
  xlab("Petal Width") +
  ylab("Petal Length")
```
